Ribosome accumulation during early phase resistance training in humans

### 2021-12-10 manuscript.Rmd
last updates before submission

---
title: "Ribosome accumulation during early phase resistance training"
author:
  - Daniel Hammarström^1,2,£^*
  - Sjur J. Øfsteng^2,£^
  - Nicolai B. Jacobsen^1^
  - Krister B. Flobergseter^1^
  - Bent R. Rønnestad^1^
  - Stian Ellefsen^1,3^
date: ''
output:
  pdf_document:
    latex_engine: xelatex
  word_document:
    reference_docx: resources/manuscript_word_template.docx
csl: resources/acta-physiologica.csl
bibliography: resources/bibliography.bib
header-includes: 
  \usepackage{wasysym}
editor_options:
  chunk_output_type: console
---
<!-- This is a comment  -->
<!-- Add comments after each reference for tracking (eg[@RN2364]<!-- comment -->
<!-- After references, add a row break without space (this will put the reference on the end on the line). -->

@@

source("./R/libs.R")
library(knitr); library(kableExtra)
# Settings for chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev = "cairo_pdf")

@@

Daniel Hammarström, Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway. Email: daniel.hammarstrom@inn.no  </div>

@@

Increased ribosomal density is a phenomenon observed in response to resistance training (RT) supporting elevated. A time course of RT-induced accumulation of ribosomes has not been described. We mapped markers of ribosomal biogenesis (total RNA, ribosomal RNA (rRNA), ribosomal protein S6 (rpS6), and upstream binding factor (UBF) protein content) in response to twelve RT sessions. Additionally, we investigated the effect of fluctuations in training volume on these markers and the relationship between RNA accumulation and muscle growth measured as muscle thickness (ultrasound). Eighteen participants were allocated to either a training (TRAIN, *n*=11) or a control (CTRL, *n*=7) group. In TRAIN, participants performed unilateral knee extension with either constant (CONST, 6 sets) or a variable volume (VAR, 6, 3, and 9 sets in sessions 1-4, 5-8, and 9-12, respectively). Muscle biopsies were sampled from *m vastus lateralis* in TRAIN before and 48 hrs after the first session and 48 hrs after sessions 4, 5, 8, 9, and 12, and after eight days of de-training. Control biopsies were sampled in CTRL at baseline and after 48 hrs and 3-5 weeks. RT led to muscle growth in TRAIN compared to CTRL and concomitant increases in total RNA, rRNA, including precursor rRNA as well as UBF and rpS6 protein through gradual increase throughout the training period, confirmed in comparison to CTRL. Total RNA increased in a curve-linear fashion, most rapidly in response to the first four sessions  (8.6%, 95% CI: [5.5, 11.7] per session), followed by a plateau and peak values (49.5% [34.2, 66.5] above baseline) after Session 8. UBF protein levels explained total RNA levels after controlling for time, and increases in total RNA levels predicted RT-induced muscle hypertrophy. After de-training, total RNA and specific rRNA species decreased without muscle mass changes indicating reduced concentrations and biosynthesis of ribosomes. These results underline a determinant role for ribosomal biogenesis in RT-induced muscle hypertrophy and that ribosomal biogenesis is sensitive to training cessation.

**Keywords:** Resistance training, ribosome biogenesis

@@

Based on these observations we hypothesize that (1) ribosome accumulation occurs during the early phase (3-4 weeks) of RT, within which this accumulation (2) reaches a plateau when RT volume is kept constant, (3) displays fluctuations in response to fluctuating training volume and (4) is partially reversed to one week after cessation of RT. In addition to addressing these hypotheses we aimed to relate RNA accumulation to total UBF levels and muscle growth.

# Methods

Eighteen volunteers were recruited to the study. Eligible participants were non-smokers between 18 and 35 years of age with a training history of less than one RT session per week during the six months leading up to the study. Exclusion criteria were consumption of dietary supplements or medication with known effects on muscle metabolism, injuries causing impaired strength and/or affecting their ability to perform RT, symptoms or history of disease, and known adverse reactions to local anesthetics. Participants were allocated to either an training group (TRAIN, *n* = 11) or a non-training control group (CTRL, *n* = 8; see Table 1 for participant characteristics; see Figure 1A for overview of the intervention). TRAIN performed a 12 session RT protocol lasting for 3-4 weeks, consisting of 10 repetition maximum (RM) unilateral knee-extension, with the two legs conducting RT with different volume profiles, allowing within-participant comparison of the effects of volume regimes. In TRAIN, one leg conducted RT with constant volume throughout the intervention (CONST, 6 sets per session) and the other leg performed RT with variable volume (VAR, 3 blocks of four sessions with 6, 3 and 9 sets per session, respectively; Figure 1A). CTRL did not partake in RT and were instructed to continue their everyday activities. Muscle biopsies were sampled bilaterally in TRAIN before and 48 hrs after the first session, as well as 48 hrs after the fourth, fifth, eight, ninth and twelfth session, and after a seven days of de-training. Muscle biopsies were obtained from CTRL at three occasions; at baseline and 48 hrs and 3-5 weeks (average (SD) `r ctrl`) after the first sampling event. TRAIN and CTRL performed strength assessments > seven days prior to the first biopsy sampling (TRAIN; CTRL), 72 hrs after the twelfth session (TRAIN) and 24 h after the last biopsy (TRAIN, following de-training; CTRL). Appendicular lean mass (Dual-energy X-ray absorptiometry, DXA) and muscle thickness of *m. vastus lateralis* were assessed prior to the first biopsy (TRAIN and CTRL) as wells as before the second to last (TRAIN) and last (TRAIN and CTRL) biopsy.

```{r table1}

  mutate(variable =  factor(variable,
                          levels = c("n", "age", "mass", "stature", "BMI", "fat"), 
                          labels = c("n", "Age (yrs)", "Body mass, (kg)", "Stature (cm)", "Body mass index (kg/m\\textsuperscript{2})", "Body fat (\\%)"))) %>%

  dplyr::select(variable, F_experiment, M_experiment, F_control, M_control) %>%

  kable(caption = "Participant characteristics.", 
        booktabs = TRUE, 
        col.names = c("", "Female", "Male", "Female", "Male"),
        escape = FALSE,
        format = "latex") %>%
  add_header_above(c(" ", "Experimental group" = 2, "Control group" = 2))

## Quantitative polymerase chain reaction (qPCR)
Complementary DNA (cDNA) was synthesized in technical duplicates from 500 ng of total RNA using random hexemer and anchored Oligo-dT primers (Thermo Fisher Scientific) together with Superscript IV (Thermo Fisher Scientific) according to manufacturer's instruction. qPCR reactions were performed with diluted cDNA (2 &#956;l, 1:25 dilution), a SYBR-green based commercial master mix (PowerUp™ SYBR™ Green Master Mix, Thermo Fisher) and, target-specific primers (500 nM) in 10 &#956;l reaction volumes using a real-time detection system (QuantStudio 5 Real-Time PCR System, Thermo Fisher Scientific). Fast cycling was used (1 sec denaturing, 30 sec annealing) after UNG (2 min, 50&deg;C) and polymerase  (2 min, 95&deg;C) activation. Melt curves were collected from all reactions to confirm single product amplification. Primers were further evaluated by agarose gel electrophoresis which confirmed primer sizes and non-template control experiments confirming no amplification without template. Primer sequences and their respective average performances are shown in Table 2.
```{r check-primer-sizes, include = FALSE}

```{r primerstable}
# Read qpcr data
qdat <- readRDS("./data/derivedData/qpcr/qpcr_compiled.RDS")

  dplyr::select(symbol_ext, full_name, forward, reverse, performance) %>%
  pivot_longer(names_to = "primer", values_to = "seq", forward:reverse) %>%
dplyr::select(Symbol = symbol_ext, 'Transcript name' = full_name, Sequence = seq, 'Mean Cq (SD) and efficiency' = performance) %>%
    filter(!(Symbol %in% c("RPL32", "MYH7", "MYH2", "MYH1")))

primer_table %>%
  kable(format = "latex", 
        caption = "Primer sequences and average performance.",
        booktabs = TRUE) %>%
  collapse_rows(columns = c(1, 2, 4)) %>%
  landscape()

# 
# flextable() %>%
# merge_custom(x = primer_table$Symbol, columns = c(1, 2, 4)) %>%
# add_header_lines(values = "Table 2. Primer sequences and average performance.") %>%
#     fontsize(size = 8, part = "body") %>%
#   fontsize(size = 10, part = "header") %>%
# autofit()
#

## Immunoblotting
Protein samples (20 &#956;g) were separated on 4-20% Tris-Glycin gels (Criterion TGX Precast Gels, Bio-Rad) at 250 V for 45 min using the recommended running buffer (25 mM Tris, 192 mM Glycin, 0.1% SDS). All samples from the same participant were run on the same gel and all samples were run in at least duplicates. Separated samples were transferred to &#956m PVDF membranes (Immun-Blot, Bio-Rad) using wet transfer (25 mM Tris, 192 Glycin, 10% vol/vol methanol) at a constant voltage of 300 mA for 3 h. Membranes were then stained to confirm transfer and enable total protein quantification using a reversible protein stain (Pierce Reversible Protein Stain, Thermo Fisher Scientiﬁc). Primary antibodies were acquired to detect UBF (F-9, sc-13125, Santa-Cruz Biotechnology, Dallas, Texas, USA) and ribosomal protein S6 (54D2, #2317, Cell Signaling Technology, Danvers, MA, USA). After blocking (Tris-buffered saline blocking buffer, 20 mM Tris, 150 mM NaCl, 5% fat-free milk, 0.1% Tween-20), membranes were incubated over-night with primary antibodies diluted in blocking buffer (UBF, 1:200; S6, 1:1000) followed by incubation with a secondary antibody conjugated to horseradish peroxidase (Anti-mouse IgG, #7076, Cell Signaling Technology, 1:10000). Membranes were washed 6 &#215; 5 min after incubation with primary antibodies and 8 &#215; 5 min after incubation with the secondary antibody. All incubation and washing steps were performed at 4&deg;C using an automatic membrane processor (BlotCycler, Precision Biosystems, Mansﬁeld, MA, USA). Chemiluminescent signals from membranes were detected after 5 min incubation in substrate (Super Signal West Femto Maximum Sensitivity Substrate, Thermo Fisher Scientiﬁc) using a documentation system. Total protein content was quantified from whole membrane images and defined as the mean gray value of the whole lane. Between-lane gray values were used as background subtracted from protein values. Total protein quantification was done using ImageJ Fiji [@RN2561]. Chemiluminescence signals were quantified using Image Studio Lite (LI-COR Biotechnology, Lincoln, NE, USA).

## Statistics and data analysis
Descriptive data are presented as mean and standard deviation (SD). The effect of training on muscle strength, muscle thickness, UBF/rpS6 protein, total RNA and gene abundances were assessed using mixed effects regression models. Time and group (TRAIN vs. CTRL) were treated as population (fixed) effects and leg nested within participant included as group level (random) effects. These analyses were performed on data with matching time points between TRAIN and CTRL with the exeption that all post-training data from TRAIN were included (post-training and de-training). Relative interactions between groups were estimated as $\Delta$ TRAIN - $\Delta$ CTRL. The effects of different volume conditions and general time-course patterns were assessed using all pairwise observations from the TRAIN group. For protein and total RNA data, segmented regression models were used to estimate changes over sessions in three segments (session 1-4, 4-8 and 8-12; corresponding to blocks of different volume prescription in TRAIN). When no robust effects of volume conditions were detected, group averages are presented. Segmented models were fitted with time and volume condition as population effects and legs nested within participants as group level effects.

Muscle strength, muscle thickness, protein and total RNA data was modeled after log transformation.
Gene abundance data were fitted with number of sessions as a categorical variable in comparisons of volume conditions, and Cq values converted to counts as suggested by Matz et al. [-@RN1964]. A Poisson-lognormal model was used to fit these count, using data from all genes and including group level effects for each technical duplicate, controlling for technical errors during sample preparation. An offset consisting of a normalization factor proportional to the amount of muscle used to prepare cDNA was used to model gene abundance per tissue weight. The external reference gene was used to calculate the normalization factor (External reference counts $\times$ muscle weight (mg) in each Trizol preparation). The offset was specified as a predictor with the coefficient fixed to 1.

A linear model was used to estimate the increase per session and average total RNA for every leg in the TR group. These estimates were then used to estimate the effect of training-induced increase per session and average total RNA abundance on muscle hypertrophy. For each leg, session was used as the independent variable centered on Session 6 and log transformed RNA per tissue weight as the dependent variable. Mean-centering of the independent variable was done to obtain an estimate of the average RNA concentration per leg. This also assured that the slope and intercept did not correlate, something that could lead to colinearity issues in subsequent modeling. A mixed effects model was subsequently fitted with differences in muscle thickness pre- to post-training as the dependent variable and estimated percentage per session increases in total RNA, the mean total RNA scaled as standard deviations from the mean and sex as independent variables. Leg nested within participant was used as group levels effects. The robustness of this model was assessed by leave-one-out analysis on the level of individual data points in the relationship between total RNA and sessions and on the level of participants (see Results).

All models were fitted using a Bayesian framework using either the brms [@RN2562] or MCMCglmm [@RN1992] package written for R [@Rcore]. Inference about effects of interest was done based on point estimates and their 95% credible intervals (CI). Credible intervals not containing null effects were interpreted as robust. Models were fitted with default priors which also makes CI analogous to confidence intervals but with the interpretation being that the CI contains the true population value with the specified certainty (95%), given the data.
Fitting performance was assessed by confirming convergence of at least four different chains of MCMC samples (graphically assessed and confirmed with $\hat{R}\approx 1$). Model performance was assessed from comparing simulated data from each model to observed data graphically (posterior predictive checks).

@@

load.inr.w2 <- paste0(sprintf("%.0f",h[1, 2]), 
                      "%, 95% CI: [",
                      sprintf("%.0f",h[1, 4]), 
                      ", ",
                      sprintf("%.0f",h[1, 5]), 
                      "]")
load.inr.w3 <- paste0(sprintf("%.0f",h[2, 2]), 
                      "% [",
                      sprintf("%.0f",h[2, 4]), 
                      ", ",
                      sprintf("%.0f",h[2, 5]), 
                      "]")

@@

In TRAIN, all participants successfully completed their prescribed RT on both legs, with the two volume conditions resulting in diverging volume profiles (load &#215; repetitions) over the course of the study (Figure 1B). Exercise intensities (resistance at 10RM) increased similarly in both conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) training block, with each block consisting of four training sessions.
Concomitantly, in TRAIN, isokinetic strength and thickness of *m. vastus lateralis* increased from baseline to after Session 12 compared to CTRL (isokinetic strength $\sim$ `r isok_pre_post_aprox`%-point difference; muscle thickness $\sim$ `r us_pre_post_approx`%-point difference, Figure 1C and D), a difference that was sustained to after eight days of de-training ($\sim$ `r isok_pre_detrain_approx`%-point and $\sim$ `r us_pre_detrain_approx`%-point difference in change in isokinetic strength and muscle thickness, respectively; Figure 1C and D). Isometric strength showed the same general pattern to after Session 12 ($\sim$ `r us_pre_detrain_approx`%-point difference), though with considerably larger degrees of uncertainty, as indicated by wider 95% CI normalization compared to CTRL after de-training ($\sim$ `r isom_pre_detrain_approx`%-point; Figure 1C). No differences were observed between volume conditions for either strength or muscle thickness.

```{r figure1, fig.height=(23*0.75)/2.54, fig.width=17.8/2.54, fig.align="center", fig.cap="Study design (A) observed training loads in response to Constant and Variable volume protocols (B) and training outcomes (muscle strength, C; muscle thickness D).", echo=FALSE}

figure1 <- readRDS("./Figures/rds/figure1.RDS")

figure1

@@

For both rpS6 and UBF, protein levels increased linearly throughout the training intervention, with rpS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`. This general pattern was confirmed when comparing TRAIN to CTRL where UBF and rpS6 protein levels were higher in TRAIN compared to CTRL after Session 12 and elevated after eight days of rest (Figure 2A and B), however with no differences being observed after the first training session (48 hrs). Increases did not differ between volume-conditions but for UBF, there was a tendency towards lower levels in VAR after Session 12 (`r var_s12`). After de-training, UBF-levels tended to decrease in CONST (`r const_detrain`) while levels in VAR remained at elevated level compare to after Session 12 (`r var_change`; interaction effect: `r var_const_interaction`). For rpS6, de-training affect protein levels, which remained similar between volume conditions period and remained similar between volume conditions. At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while rpS6 showed no robust differences between TRAIN and CTRL at any time point. No differences were observed between volume conditions for either transcripts (Figure 2D and E).

```{r figure2, fig.height=(23*0.75)/2.54, fig.width=8.9/2.54, fig.align="center", fig.cap="Protein (A and B) and mRNA abundances (D and E) of rpS6 and UBF. Black points and error bars represents statistically robust results (a 95\\% CI not containing 0). C shows western-blots and total protein stains from a representative participant. mRNA data is normalized per total RNA.", echo=FALSE}

figure2 <- readRDS("./Figures/rds/figure2.RDS")

figure2

@@

A single session of RT (Session 1) led to robust increases in pre-rRNA 47S ETS and 45S ETS abundance per unit tissue weight, measured as changes from baseline to 48 hrs after exercise within TRAIN (Figure 3B), as well as compared to CTRL (Figure 3C).
After Session 1, pre-rRNA 47S ETS and 45S ETS levels remained at similar levels at all measured time-points in TRAIN (Figure 3D), confirmed in comparison to CTRL after Session 12 (Figure 3C).
Other rRNA transcripts showed increases in response to training with slightly different temporal patterns with exception of rRNA 5S which did not change and rRNA 5.8 which tended to follow other mature transcript spliced from pre-rRNA 45S ETS, but without statistical robustness (Figure 3C and D).
After eight days of rest, 18S and 28S remained at elevated levels compared to CTRL (Figure 3B).
This general pattern of rRNA expression was reflected by total RNA expression per unit tissue weight, which increased robustly and steadily in TRAIN throughout the initial part of the intervention (Figure 3E and G), leading to robust increase compared to CTRL after Session 12 (Figure 3F), followed by decreased levels after de-training (`r detrain_rna`).
For both rRNA expression and total RNA levels, the training-associated increases in abundances occurred predominately during the first four sessions, evident as `r slope0_4` increase per session, followed by sustained levels from sessions four to eight (`r slope4_8` increase per session) and from sessions eight to twelve `r slope8_12`, corresponding to `r fourth.session`, `r eight.session` and `r twelth.session` increases from baseline to 48 hrs after session four, eight and twelve, respectively.
In TRAIN, the two volume conditions led to similar changes for most variables (Figure 3D and G), with 45S ETS abundance only showing differential expression, evident as robustly higher levels in VAR compared to CONST after the 12th session (Figure 3D), coinciding with the increased training volume towards the end of the intervention for this condition.

```{r figure3, fig.height=(23*0.75)/2.54, fig.width=17.8/2.54, fig.align="center", fig.cap="Total RNA and ribosomal RNA subspecies in response to trainin. Total RNA increased compared to non-training controls to post-training (12 sessions) and tended to normalize after de-training (A). There were no robust differences between different volume conditions in the training group (A). B shows primer locations targeting different ribosomal RNA subspecies. Subsets of ribosomal RNA showed robust increases compared to control 48 h after the first session and other after the training period (C). Error bars shows 95\\% CI. Asterisk in C indicates robust differences between volume conditions (a 95\\% CI of pairwise differences not containing 0).", echo=FALSE}

figure3 <- readRDS("./Figures/rds/figure3.RDS")

figure3

```{r table3-regression-models-muscle-thickness}
# Save model

reg_results <- 
  cbind(rbind(comb.m1$fixed,
      comb.m1$random$participant, 
      comb.m1$random$`participant:leg`, 
      comb.m1$spec_pars), 
      data.frame(coef = c("Intercept", 
                          "UBF protein levels (SD-units)", 
                          "Session 1-4\\textsuperscript{b}",
                          "Session 4-8\\textsuperscript{c}", 
                          "Session 8-12\\textsuperscript{d}", 
                          "De-training",
                          "Between Participant variation",
                          "Between Participant:Leg variation", 
                          "Residual SD"))) %>%
  remove_rownames() %>%
  dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)

reg_results_s6 <- 
  cbind(rbind(comb.s6.m1$fixed,
      comb.s6.m1$random$participant, 
      comb.s6.m1$random$`participant:leg`, 
      comb.s6.m1$spec_pars), 
      data.frame(coef = c("Intercept", 
                          "rpS6 protein levels (SD-units)", 
                          "Session 1-4\\textsuperscript{b}",
                          "Session 4-8\\textsuperscript{c}", 
                          "Session 8-12\\textsuperscript{d}", 
                          "De-training",
                          "Between Participant variation",
                          "Between Participant:Leg variation", 
                          "Residual SD"))) %>%
  remove_rownames() %>%
  dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)

    kable(col.names = c("Coefficient", 
                        "Estimate\\textsuperscript{a}", 
                        "SD", 
                        "Lower 95\\% CI", 
                        "Upper 95\\% CI"), 
          caption = "Effect of UBF and rpS6 levels, sessions and de-training on RNA-levels.",
          format = "latex", 
          escape = FALSE,
          booktabs = TRUE) %>%
    row_spec(9, hline_after = TRUE) %>%
    footnote(alphabet = c("The dependent variable is total RNA levels (log).",
    "Slope in response to session 1-4.", 
    "Change in slope in session 4-8.",
    "Change in slope in session 8-12."), 
             footnote_as_chunk = TRUE,
             threeparttable = TRUE, 
             escape = FALSE,
             general_title = "")

@@

```{r table4-regression-models-muscle-thickness}
# Save model

>>>>

---
title: "Ribosome accumulation during early phase resistance training in humans"
author: Daniel Hammarström^1,2,£^*, Sjur J. Øfsteng^2,£^*, Nicolai B. Jacobsen^1^, Krister B. Flobergseter^1^, Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
date: ''
output:
  officedown::rdocx_document:
    reference_docx: resources/manuscript_word_template.docx
csl: resources/acta-physiologica.csl
bibliography: resources/bibliography.bib
header-includes: 
  \usepackage{wasysym}
editor_options:
  chunk_output_type: console
---
<!-- This is a comment  -->
<!-- Add comments after each reference for tracking (eg[@RN2364]<!-- comment -->
<!-- After references, add a row break without space (this will put the reference on the end on the line). -->

@@

source("./R/libs.R")
library(knitr); library(kableExtra); library(officedown)
# Settings for chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev = "cairo_pdf")

@@

Daniel Hammarström, Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway. Email: daniel.hammarstrom@inn.no

Sjur Johansen Øfsteng, Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway. Email: sjur.johansen.ofsteng@inn.no
</div>

@@

```{r}

### Data in abstract ###########

# Ultra sound #################

# Model data
comp_us <- readRDS("./data/derivedData/us-freq-bayes/comp_us_bayes.RDS") %>%
   dplyr::select(time_group, estimate, lower.CL, upper.CL)

# Express as percentage of average
library(tidyverse); library(readxl)

us_data <- rbind(read_excel("./data/ultrasound/ultrasound_data.xlsx") %>%
                   inner_join(read_csv("./data/ultrasound/ultrasound_codekey.csv")) %>%
                   mutate(leg = gsub("VL", "", leg)) %>%
                   inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
                   dplyr::select(participant, time, leg, sex, cond, code, length) %>%
                   mutate(group = if_else(participant %in% paste("P", 1:7, sep = ""), "experiment", "control")) %>%
                   group_by(participant, time, leg, sex, cond, group) %>%
                   summarise(thickness = mean(length, na.rm = TRUE)) %>%
                   ungroup(), 
                 read_excel("./data/ultrasound/ultrasound_data_2019.xlsx") %>%
                   inner_join(read_csv("./data/ultrasound/ultrasound_codekey_2019.csv")) %>%
                   mutate(leg = gsub("VL", "", leg)) %>%
                   inner_join(read_excel("./data/leg_randomization.xlsx")) %>%
                   dplyr::select(participant, time, leg, sex, cond, code, length) %>%
                   mutate(group = if_else(participant %in% paste("P", c(1:7, 19:23), sep = ""), "experiment", "control")) %>%
                   group_by(participant, time, leg, sex, cond, group) %>%
                   summarise(thickness = mean(length, na.rm = TRUE)) %>%
                   ungroup()) %>%

  mutate(time_pp = if_else(time == "post1w", "post", time),
         # The de-training period get its own coefficient
         detrain = if_else(time == "post1w" & group == "experiment", "detrain", "train"),
         # The effect of de training will be added to the model --within-- the intervention group 
         detrain = factor(detrain, levels = c("train", "detrain")), 
         time = factor(time, levels = c("pre", "post", "post1w")), 
         time_pp = factor(time_pp, levels = c("pre", "post"))) %>%
  group_by(time) %>%
  summarise(m = mean(thickness, na.rm = TRUE)) %>%
  data.frame()

hypertrophy <- paste0(sprintf("%.1f", 100 * (comp_us[4,2] / us_data[1,2])), 
       "% over baseline values, 95% CrI: [",
       sprintf("%.1f", 100 * (comp_us[4,3] / us_data[1,2])), 
       ", ",
       sprintf("%.1f", 100 * (comp_us[4,4] / us_data[1,2])),
       "]" )

## Total RNA increase ######################

# Segmented model analysis in text 
# Read the segmented model

library(brms)

tc.m2 <- readRDS("./data/derivedData/total-rna-analysis/tc.m2.RDS")

time1.hyp <- hypothesis(tc.m2, "time1 = 0")

time2.hyp <- hypothesis(tc.m2, c("time2 + time1 = 0")) # The slope (on log scale) between session 4 and 8
time3.hyp <- hypothesis(tc.m2, c("time2 + time1 +time3 = 0")) # The slope on log scale between session 8 and 12

detrain_rna <- paste0(sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,2]) - 1)),
                           "% [",
                      sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,4]) - 1)), 
                   ", ",
                       sprintf("%.1f", 100 * (exp(detrain.hyp$hypothesis[1,5]) - 1)),
                   "]")

# UBF effect on total RNA ############
comb.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1.RDS")

ubf_model <- paste0(sprintf("%.1f", (exp(summary(comb.m1)$fixed[2,1])-1) * 100), 
       "% [",
       sprintf("%.1f", (exp(summary(comb.m1)$fixed[2,3])-1) * 100), 
       ", ",
       sprintf("%.1f", (exp(summary(comb.m1)$fixed[2,4])-1) * 100),
       "]" )

# Rate of increase in total rna on muscle hypertrophy ################

hyp.pre.m3 <- readRDS("./data/derivedData/ubf-tot-rna-model/hyp_pre_m3.RDS")

rna_hypertrophy <- paste0(sprintf("%.1f", summary(hyp.pre.m3)$fixed[5,1]), 
       " mm [",
       sprintf("%.1f", summary(hyp.pre.m3)$fixed[5,3]), 
       ", ",
       sprintf("%.1f", summary(hyp.pre.m3)$fixed[5,4]),
       "]" )

**Aim:** To describe the time course of ribosome accumulation during resistance training, its relation to training volume, UBF protein levels, and muscle growth.

**Methods:** A training group (n=11) performed unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg. Vastus lateralis biopsies obtained at rest and 48 hrs after selected training sessions and after eight days of de-training and from a non-training control group (n=7) were assayed for ribosome abundance and biogenesis markers. In addition, muscle thickness was measured before and after training and control periods.

**Results:** Training led to muscle growth (`r hypertrophy` vs. control) with concomitant increases in total RNA, rRNA, UBF, and rpS6 with no differences between volume conditions. Total RNA increased most rapidly in response to the first four sessions (`r slope0_4` per session), followed by a plateau and peak values after session 8 (`r eight.session` above baseline). After controlling for time, total RNA abundance was associated with UBF protein levels (`r ubf_model` per unit UBF), and the rate of increase in total RNA levels predicted hypertrophy (`r rna_hypertrophy` per %-point increase in total RNA per session). After de-training, total RNA decreased (`r detrain_rna`) without muscle mass changes indicating halted biosynthesis of ribosomes.

**Conclusion:** Ribosomes accumulate in the initial phase of resistance training with abundances sensitive to training cessation and associated with UBF protein levels. The average accumulation rate predicts muscle training-induced hypertrophy.

**Keywords:** muscle hypertrophy, resistance training, ribosome biogenesis

@@

Based on these observations we hypothesize that (1) ribosome accumulation occurs during the early phase (3-4 weeks) of RT, within which this accumulation (2) reaches a plateau when RT volume is kept constant, (3) displays fluctuations in response to fluctuating training volume and (4) is partially reversed to one week after cessation of RT. In addition to addressing these hypotheses we aimed to relate RNA accumulation to total UBF levels and muscle growth. We utilized a within-participant unilateral training model where one leg was assigned constant volume (CONST, 6 sets per session) and the contra lateral leg variable volume (VAR, 6, 3, and 9 sets in sessions 1-4, 5-8, and 9-12, respectively). Effects of training *per se* were assessed by comparison to a non-training control group (CTRL).

@@

load.inr.w2 <- paste0(sprintf("%.0f",h[1, 2]), 
                      "%, 95% credible interval (CI): [",
                      sprintf("%.0f",h[1, 4]), 
                      ", ",
                      sprintf("%.0f",h[1, 5]), 
                      "]")
load.inr.w3 <- paste0(sprintf("%.0f",h[2, 2]), 
                      "% [",
                      sprintf("%.0f",h[2, 4]), 
                      ", ",
                      sprintf("%.0f",h[2, 5]), 
                      "]")

@@

All participants allocated to TRAIN successfully completed their prescribed RT on both legs, with the two volume conditions resulting in diverging volume profiles (load &#215; repetitions) over the course of the study (Figure 1B). Exercise intensities (resistance at 10RM) increased similarly in both conditions from the first to the second (`r load.inr.w2`) and third (`r load.inr.w3`) training block, with each block consisting of four training sessions.
Concomitantly, in TRAIN, isokinetic strength and thickness of *m. vastus lateralis* increased from baseline to after Session 12 compared to CTRL (isokinetic strength $\sim$ `r isok_pre_post_aprox`%-point difference; muscle thickness $\sim$ `r us_pre_post_approx`%-point difference, Figure 1C and D), a difference that was sustained to after eight days of de-training ($\sim$ `r isok_pre_detrain_approx`%-point and $\sim$ `r us_pre_detrain_approx`%-point difference in change in isokinetic strength and muscle thickness, respectively; Figure 1C and D). Isometric strength showed the same general pattern to after Session 12 ($\sim$ `r us_pre_detrain_approx`%-point difference), though with considerably larger degrees of uncertainty, as indicated by wider 95% CI normalization compared to CTRL after de-training ($\sim$ `r isom_pre_detrain_approx`%-point; Figure 1C). No differences were observed between volume conditions for either strength or muscle thickness.

@@

For both ribosomal protein S6 (rpS6) and upstream binding factor (UBF), protein levels increased linearly throughout the training intervention, with rpS6 showing estimated increments per session corresponding to `r s6_slope0_4` during block 1 (session 1-4), `r s6_slope4_8` during block 2 (session 5-8) and `r s6_slope8_12` during block 3 (session 9-12), and UBF showing increments corresponding to `r ubf_slope0_4`, `r ubf_slope4_8` and `r ubf_slope8_12`. This general pattern was confirmed when comparing TRAIN to CTRL where UBF and rpS6 protein levels were higher in TRAIN compared to CTRL after Session 12 and elevated after eight days of rest (Figure 2A and B), however with no robust differences being observed after the first training session (48 hrs). Increases did not differ between volume-conditions but for UBF, there was a tendency towards lower levels in VAR after Session 12 (`r var_s12`). After de-training, UBF-levels tended to decrease in CONST (`r const_detrain`) while levels in VAR remained at elevated level compare to after Session 12 (`r var_change`; interaction effect: `r var_const_interaction`). For rpS6, de-training did not affect protein levels, which remained similar between volume conditions. At the mRNA level, UBF showed robust increase from before to 48 hrs after the first session in TRAIN compared to CTRL (Figure 2D), while rpS6 showed no robust differences between TRAIN and CTRL at any time point. No differences were observed between volume conditions for either transcripts (Figure 2D and E).

@@

A single session of RT (Session 1) led to robust increases in precursor ribosomal RNA (pre-rRNA 47S ETS and 45S ETS) abundance per unit tissue weight, measured as changes from baseline to 48 hrs after exercise within TRAIN (Figure 3B), as well as compared to CTRL (Figure 3C).
After Session 1, pre-rRNA 47S ETS and 45S ETS levels remained at similar levels at all measured time-points in TRAIN (Figure 3D), confirmed in comparison to CTRL after Session 12 (Figure 3C).
Other rRNA transcripts showed increases in response to training with slightly different temporal patterns with exception of rRNA 5S which did not change and rRNA 5.8 which tended to follow other mature transcript spliced from pre-rRNA 45S ETS, but without statistical robustness (Figure 3C and D).
After eight days of rest, 18S and 28S remained at elevated levels compared to CTRL (Figure 3B).
This general pattern of rRNA expression was reflected by total RNA abundance per unit tissue weight, which increased robustly and steadily in TRAIN throughout the initial part of the intervention (Figure 3E and G), leading to robust increase compared to CTRL after Session 12 (Figure 3F), followed by decreased levels after de-training (`r detrain_rna`).
For both rRNA expression and total RNA levels, the training-associated increases in abundances occurred predominately during the first four sessions, evident as `r slope0_4` increase per session, followed by sustained levels from sessions four to eight (`r slope4_8` increase per session) and from sessions eight to twelve `r slope8_12`, corresponding to `r fourth.session`, `r eight.session` and `r twelth.session` increases from baseline to 48 hrs after session four, eight and twelve, respectively.
In TRAIN, the two volume conditions led to similar changes for most variables (Figure 3D and G), with 45S ETS abundance only showing differential expression, evident as robustly higher levels in VAR compared to CONST after the 12th session (Figure 3D), coinciding with the increased training volume towards the end of the intervention for this condition.

@@

In TRAIN, total RNA levels were robustly predicted by UBF levels (after controlling for time), with `r ubf_cor_tot_rna` increases in total RNA per unit tissue weight coinciding with one unit increase in UBF levels (corresponding to one standard deviation; Table 1). In contrast, no evidence was found for a relationship between total RNA and rpS6 protein levels (Table 1).

In TRAIN, there was a robust positive relationship between rates of increase in total RNA in response to training and muscle growth measured as increases in *m. vastus lateralis* thickness (Table 2, Figure 4A), with changes in total RNA over the course of the training intervention being estimated in each leg using a regression model containing number of sessions as the independent variable. Conversely, there was a tendency towards a negative relationship between average total RNA levels at Session 6 and changes in muscle thickness (Table 2, Figure 4B), with the average total RNA levels estimated as the predicted value at Session 6 (estimated as the intercept-term) from the model used to estimate the rate of total RNA increase per session.
